Abstract Binding Trees

### 2015-11-07 paper.tex
[paper] notation changes to comport with Agda development
let σ, τ range over sorts; use fraktur for metavariables

\newcommand\MV[1]{\textsc{#1}}
\newcommand\MApp[3]{#1\{#2\}(#3)}
\newcommand\App[2]{#1(#2)}
\newcommand\Dom[1]{\left\vert#1\right\vert}
\newcommand\FV[1]{\mathbf{FV}\left(#1\right)}
\newcommand\FS[1]{\mathbf{FS}\left(#1\right)}
\newcommand\NotIn[2]{\IMode{#1}\notin\IMode{#2}}
\newcommand\SortExp{\mathsf{exp}}

@@

Fix a set $\Sorts$ of \emph{sorts}. We will say $\IsSort{s}$ when
$\Member{s}{S}$. A valence $\MkValence{\vec{p}}{\vec{q}}{s}$ specifies an
expression of sort $s$ which binds symbols in $\vec{p}$ and variables in
$\vec{q}$.

\[
  \infer{
    \IsValence{\MkValence{p_0,\dots,p_m}{q_0,\dots,q_n}{s}}
  }{
    \IsSort{s} &
    \HypJ{\IsSort{p_i}}{\Leq{i}{m}} &
    \HypJ{\IsSort{q_i}}{\Leq{i}{n}} &
  }
\]

An arity $\MkArity{\vec{v}}{s}$ specifies an operator of sort $s$ with
arguments of valences $\vec{v}$. We will call the set of valences $\Valences$,
and the set of arities $\Arities$.

\[
  \infer{
    \IsArity{\MkArity{v_0,\ldots,v_n}{s}}
  }{
    \IsSort{s} &
    \HypJ{\IsValence{v_i}}{\Leq{i}{n}}
  }
\]

@@

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsVarCtx{\Gamma,x:s}
  }{
    \IsVarCtx{\Gamma} &
    \IsSort{s} &
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:s}
  }{
    \IsVarCtx{\Upsilon} &
    \IsSort{s} &
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{s}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{s}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees (abts) is inductively defined in four rules:

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{s}
  }{
    \Lookup{\Gamma}{x}{s}
  }\\[6pt]
  %
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{s}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{p_0,\dots,p_m}{q_0,\dots,q_n}{s}}\\
      \HypJ{\Lookup{\Upsilon}{u_i}{p_i}}{\Leq{i}{m}}\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  %
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{s}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        v_1,\dots,v_n
      }\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{p}}{\vec{q}}{s}
    }
  }{
    \IsAbt{\Omega}{\Upsilon, \vec{u}:\vec{p}}{\Gamma,\vec{x}:{\vec{q}}}{
      M
    }{
      s
    }
  }
\end{gather*}

@@

The only place that symbols appear in our calculus is as parameters to
operators (unlike variables, symbols are not terms). Therefore, the functorial
action of the operator presheaf can be lifted into terms by recursion on their
structure, via a pair of judgments $\IsRenaming{\varrho}{M}{N}$ and
$\IsRenaming{\varrho}{E}{F}$, presupposing
$\Of{\varrho}{\Upsilon\to\Upsilon'}$, and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{s}$
and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$ respectively:

@@

Let $\Define{V_s(\Upsilon\parallel\Gamma)}{\MkSet{x\in\Dom\Gamma\mid
\Gamma(x)=s}}$ be called the presheaf of variables; additionally, we have a
presheaf of symbols
$\Define{S_s(\Upsilon\parallel\Gamma)}{\MkSet{u\in\Dom\Upsilon\mid
\Upsilon(x)=s}}$. Lastly, we have the presheaf of operators with arity $a$,
$\Define{\Operators_a(\Upsilon\parallel\Gamma)}{\Operators\Pair{\Upsilon}{a}}$

@@

Using this, we can define a tensor $P\odot Q$ for
$\Of{P,Q}{\widehat{\HypCat}^\Sorts}$ as follows:
\[
  \HypJ{
    \Define{{(P\odot Q)}_{s}}{
      P_s\bullet Q
    }
    \quad
  }{
    \Member{s}{\Sorts}
  }
\]

@@

%%TODO
We also have the induced maps $\Of{\nu_\Gamma}{V^{[\Gamma]}\to P^{[\Gamma]}}$
and
$\Of{\varsigma_{\Upsilon\parallel\Gamma}}{P_s^{\Yoneda{\Upsilon\parallel\Gamma}} \times
S^{\{\Upsilon\}}\times P^{[\Gamma]}\to P_s}$.

@@

For each signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we have an
endofunctor
$\Of{\SigFun{\Sigma}}{\widehat{\HypCat}^\Sorts\to\widehat{\HypCat}^\Sorts}$,
which is defined as follows:
\begin{align*}
  \IMode{{\SigFun{\Sigma}(X)}_{s}}
    &\triangleq
    \OMode{
      \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{s}}}
      \prod_{\MkValence{\vec{p}}{\vec{q}}{s'}\in\vec{v}}
      X_{s'}^{\Yoneda{\vec{p}\parallel\vec{q}}}
    }
\end{align*}

@@

The metavariable, symbol and variable contexts are interpreted in a model $P$ as an
environment presheaf in the following way:
\[
  \Define{
    \SemBrackets{
      \Omega\triangleright\Upsilon\parallel\Gamma
    }_P
  }{
    \left(\prod_{(m:\MkValence{\vec{p}}{\vec{q}}{s})\in\Omega}P_{s}^{\Yoneda{\vec{p}\parallel\vec{q}}}\right)
    \times
    S^{\{\Upsilon\}}
    \times
    V^{[\Gamma]}
  }
\]

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:s}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\to P_s}%
\]%
The interpretation of a variable
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x_j:s}_P$ simply
projects it of the environment and embeds it into the model, as follows:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_3}"]
    \& {\IMode{V^{[\Gamma]}}} \arrow[r, "\IMode{\nu_{\Gamma}}"]
    \& {\IMode{P^{[\Gamma]}}} \arrow[r, "\IMode{\pi_j}"]
    \& \IMode{P_s}
  \end{tikzcd}
  \tag{var}
\]

Next we have the interpretation of metavariable applications
\[%
  \SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
    \MApp{\MV{m}_i}{\vec{u}}{\vec{M}} : s}_P%
  \qquad(\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{p}}{\vec{q}}{s}})
\]
which proceeds by projecting the metavariable's interpretation from the
environment and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_i\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{s'}^{\Yoneda{\vec{p}\parallel\vec{q}}}\times S^{\{\vec{p}\}} \times P^{[\vec{q}]}}}
          \arrow[r, "\IMode{\varsigma_{\vec{p}\parallel\vec{q}}}"]
    \& \IMode{P_s}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{align*}
  \ADefine{\phi}{
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {S^{\{\Upsilon\}}} \arrow[r, "\Tuple{\pi_u}_{u\in\vec{u}}"]
    \& {S^{\{\vec{p}\}}}
  \end{tikzcd}
  }\\
  \ADefine{\psi}{
   \begin{tikzcd}[ampersand replacement = \&, sep=28ex]
     {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}
        \arrow[r, "\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:q}_P}_{(M,q)\in \vec{M}\otimes\vec{q}}"]
     \& {P^{[\vec{q}]}}
   \end{tikzcd}
  }
\end{align*}

@@

The above sequent has free symbols, but we can close over them by adding a form
of parametric higher-order judgment to our object language, indexed by a
collection of sorts $\vec{s}$:
\[
  \IsOperator{\Upsilon}{\OpNabla{\vec{s}}}{
    \MkArity{
      \MkValence{\vec{s}}{}{\SortJdg}
    }{\SortJdg}
  }
\]

@@

Having specified an approximation of the grammar of telescopes and sequents in
the abt logical framework, we can proceed to define proper wellformedness via
\emph{inductive refinement}. The basic idea is to introduce a new form of
(meta)-judgment $\IsWf{\Omega}{\Upsilon}{\Gamma}{M}{s}$ which expresses the
extrinsic wellformedness properties we wish to verify, presupposing
$\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{s}$. Additionally, we introduce an
analogous judgment on bound terms, $\IsWf{\Omega}{\Upsilon}{\Gamma}{E}{v}$
presupposing $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$, defined uniformly as
follows:
\[
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{p}}{\vec{q}}{s}
    }
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{p}}{\Gamma,\vec{x}:\vec{q}}{M}{s}
  }
\]

Likewise, wellformedness for variables and metavariables is defined uniformly:
\begin{gather*}
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{x}{s}
  }{
  }\qquad
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{s}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{\vec{p}}{q_0,\dots,q_n}{s}}\\
      \HypJ{
        \IsWf{\Omega}{\Upsilon}{\Gamma}{M_i}{q_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }
\end{gather*}

\begin{remark}
  Note that the refinement for variables $x$ is not trivial, since it is only
  defined in case the presupposition $\IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{s}$
  is satisfied.
\end{remark}

@@

\[
  \infer[
    \text{for $\Member{s}{\MkSet{\SortExp,\SortType}}$}
  ]{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{s}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{
          v_0,\dots,v_n
        }{s}
      }\\
      \HypJ{
        \IsWf{\Omega}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }
\]

The refinements for parametric judgment and sequents simply delegate to their
subterms as well:
\begin{gather*}
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \App{\OpNabla{\vec{s}}}{
        \MkBTm{\vec{u}}{}{J}
      }
    }{\SortJdg}
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{s}}{\Gamma}{J}{\SortJdg}
  }\\[6pt]
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      H\gg A
    }{\SortJdg}
  }{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{H}{\SortTele} &
    \IsWf{\Omega}{\Upsilon}{\Gamma}{A}{\SortType}
  }
\end{gather*}

>>>>

\newcommand\MV[1]{\mathfrak{#1}}
\newcommand\MApp[3]{#1\{#2\}(#3)}
\newcommand\App[2]{#1(#2)}
\newcommand\Dom[1]{\left\vert#1\right\vert}
\newcommand\FV[1]{\mathbf{FV}\left(#1\right)}
\newcommand\FS[1]{\mathbf{FS}\left(#1\right)}
\newcommand\NotIn[2]{\IMode{#1}\notin\IMode{#2}}
\newcommand\SortExp{\mathsf{exp}}

@@

Fix a set $\Sorts$ of \emph{sorts}. We will say $\IsSort{\tau}$ when
$\Member{\tau}{\Sorts}$. A valence $\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}$ specifies an
expression of sort $\tau$ which binds symbols in $\vec{\sigma}$ and variables in
$\vec{\tau}$.

\[
  \infer{
    \IsValence{\MkValence{\sigma_0,\dots,\sigma_m}{\tau_0,\dots,\tau_n}{\tau}}
  }{
    \IsSort{\tau} &
    \HypJ{\IsSort{\sigma_i}}{\Leq{i}{m}} &
    \HypJ{\IsSort{\tau_i}}{\Leq{i}{n}} &
  }
\]

An arity $\MkArity{\vec{v}}{\tau}$ specifies an operator of sort $\tau$ with
arguments of valences $\vec{v}$. We will call the set of valences $\Valences$,
and the set of arities $\Arities$.

\[
  \infer{
    \IsArity{\MkArity{v_0,\ldots,v_n}{\tau}}
  }{
    \IsSort{\tau} &
    \HypJ{\IsValence{v_i}}{\Leq{i}{n}}
  }
\]

@@

\begin{gather*}
  \infer{
    \IsVarCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsVarCtx{\Gamma,x:\tau}
  }{
    \IsVarCtx{\Gamma} &
    \IsSort{\tau} &
    \NotIn{x}{\Dom\Gamma}
  }
\end{gather*}

\begin{gather*}
  \infer{
    \IsSymCtx{\cdot}
  }{
  }\qquad
  \infer{
    \IsSymCtx{\Upsilon,u:\tau}
  }{
    \IsVarCtx{\Upsilon} &
    \IsSort{\tau} &
    \NotIn{u}{\Dom\Upsilon}
  }
\end{gather*}

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{\tau}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees (abts) is inductively defined in four rules:

\begin{gather*}
  \infer[\textit{var}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{\tau}
  }{
    \Lookup{\Gamma}{x}{\tau}
  }\\[6pt]
  %
  \infer[\textit{mvar}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{u_0,\dots,u_m}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{\sigma_0,\dots,\sigma_m}{\tau_0,\dots,\tau_n}{\tau}}\\
      \HypJ{\Lookup{\Upsilon}{u_i}{\sigma_i}}{\Leq{i}{m}}\\
      \HypJ{
        \IsAbt{\Omega}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  %
  \infer[\textit{app}]{
    \IsAbt{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{v_1,\dots,v_n}{\tau}
      }\\
      \HypJ{
        \IsBTm{\Omega}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }\\[6pt]
  \infer[\textit{abs}]{
    \IsBTm{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsAbt{\Omega}{\Upsilon, \vec{u}:\vec{\sigma}}{\Gamma,\vec{x}:{\vec{\tau}}}{
      M
    }{
      \tau
    }
  }
\end{gather*}

@@

The only place that symbols appear in our calculus is as parameters to
operators (unlike variables, symbols are not terms). Therefore, the functorial
action of the operator presheaf can be lifted into terms by recursion on their
structure, via a pair of judgments $\IsRenaming{\varrho}{M}{N}$ and
$\IsRenaming{\varrho}{E}{F}$, presupposing
$\Of{\varrho}{\Upsilon\to\Upsilon'}$, and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$
and $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$ respectively:

@@

Let $\Define{V_\tau(\Upsilon\parallel\Gamma)}{\MkSet{x\in\Dom\Gamma\mid
\Gamma(x)=\tau}}$ be called the presheaf of variables; additionally, we have a
presheaf of symbols
$\Define{S_\tau(\Upsilon\parallel\Gamma)}{\MkSet{u\in\Dom\Upsilon\mid
\Upsilon(x)=\tau}}$. Lastly, we have the presheaf of operators with arity $a$,
$\Define{\Operators_a(\Upsilon\parallel\Gamma)}{\Operators\Pair{\Upsilon}{a}}$

@@

Using this, we can define a tensor $P\odot Q$ for
$\Of{P,Q}{\widehat{\HypCat}^\Sorts}$ as follows:
\[
  \HypJ{
    \Define{{(P\odot Q)}_{\tau}}{
      P_\tau\bullet Q
    }
    \quad
  }{
    \Member{\tau}{\Sorts}
  }
\]

@@

%%TODO
We also have the induced maps $\Of{\nu_\Gamma}{V^{[\Gamma]}\to P^{[\Gamma]}}$
and
$\Of{\varsigma_{\Upsilon\parallel\Gamma}}{P_\tau^{\Yoneda{\Upsilon\parallel\Gamma}}
\times S^{\{\Upsilon\}}\times P^{[\Gamma]}\to P_\tau}$.

@@

For each signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we have an
endofunctor
$\Of{\SigFun{\Sigma}}{\widehat{\HypCat}^\Sorts\to\widehat{\HypCat}^\Sorts}$,
which is defined as follows:
\begin{align*}
  \IMode{{\SigFun{\Sigma}(X)}_{\tau}}
    &\triangleq
    \OMode{
      \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}
      \prod_{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau'}\in\vec{v}}
      X_{\tau'}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}
    }
\end{align*}

@@

The metavariable, symbol and variable contexts are interpreted in a model $P$ as an
environment presheaf in the following way:
\[
  \Define{
    \SemBrackets{
      \Omega\triangleright\Upsilon\parallel\Gamma
    }_P
  }{
    \left(\prod_{(m:\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau})\in\Omega}P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\right)
    \times
    S^{\{\Upsilon\}}
    \times
    V^{[\Gamma]}
  }
\]

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:\tau}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\to P_\tau}%
\]%
The interpretation of a variable
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ simply
projects it of the environment and embeds it into the model, as follows:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_3}"]
    \& {\IMode{V^{[\Gamma]}}} \arrow[r, "\IMode{\nu_{\Gamma}}"]
    \& {\IMode{P^{[\Gamma]}}} \arrow[r, "\IMode{\pi_x}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{var}
\]

Next we have the interpretation of metavariable applications
\[%
  \SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
    \MApp{\MV{m}}{\vec{u}}{\vec{M}} : \tau}_P%
  \qquad(\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}})
\]
which proceeds by projecting the metavariable's interpretation from the
environment and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \& \IMode{P_s}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{align*}
  \ADefine{\phi}{
  \begin{tikzcd}[ampersand replacement = \&, sep=huge]
    {S^{\{\Upsilon\}}} \arrow[r, "\Tuple{\pi_u}_{u\in\vec{u}}"]
    \& {S^{\{\vec{\sigma}\}}}
  \end{tikzcd}
  }\\
  \ADefine{\psi}{
   \begin{tikzcd}[ampersand replacement = \&, sep=28ex]
     {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}
        \arrow[r, "\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:q}_P}_{(M,q)\in \vec{M}\otimes\vec{\tau}}"]
     \& {P^{[\vec{\tau}]}}
   \end{tikzcd}
  }
\end{align*}

@@

The above sequent has free symbols, but we can close over them by adding a form
of parametric higher-order judgment to our object language, indexed by a
collection of sorts $\vec{\sigma}$:
\[
  \IsOperator{\Upsilon}{\OpNabla{\vec{\sigma}}}{
    \MkArity{
      \MkValence{\vec{\sigma}}{}{\SortJdg}
    }{\SortJdg}
  }
\]

@@

Having specified an approximation of the grammar of telescopes and sequents in
the abt logical framework, we can proceed to define proper wellformedness via
\emph{inductive refinement}. The basic idea is to introduce a new form of
(meta)-judgment $\IsWf{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ which expresses the
extrinsic wellformedness properties we wish to verify, presupposing
$\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$. Additionally, we introduce an
analogous judgment on bound terms, $\IsWf{\Omega}{\Upsilon}{\Gamma}{E}{v}$
presupposing $\IsAbt{\Omega}{\Upsilon}{\Gamma}{E}{v}$, defined uniformly as
follows:
\[
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}
    }
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{\sigma}}{\Gamma,\vec{x}:\vec{\tau}}{M}{\tau}
  }
\]

Likewise, wellformedness for variables and metavariables is defined uniformly:
\begin{gather*}
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{x}{\tau}
  }{
  }\qquad
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \MApp{\MV{m}}{\vec{u}}{M_0,\dots,M_n}
    }{\tau}
  }{
    \begin{array}{l}
      \Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\tau_0,\dots,\tau_n}{\tau}}\\
      \HypJ{
        \IsWf{\Omega}{\Upsilon}{\Gamma}{M_i}{\tau_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }
\end{gather*}

\begin{remark}
  Note that the refinement for variables $x$ is not trivial, since it is only
  defined in case the presupposition $\IsAbt{\Omega}{\Upsilon}{\Gamma}{x}{\tau}$
  is satisfied.
\end{remark}

@@

\[
  \infer[
    \text{for $\Member{\tau}{\MkSet{\SortExp,\SortType}}$}
  ]{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \App{\vartheta}{E_0,\dots,E_n}
    }{\tau}
  }{
    \begin{array}{l}
      \IsOperator{\Upsilon}{\vartheta}{
        \MkArity{
          v_0,\dots,v_n
        }{\tau}
      }\\
      \HypJ{
        \IsWf{\Omega}{\Upsilon}{\Gamma}{E_i}{v_i}
      }{
        \Leq{i}{n}
      }
    \end{array}
  }
\]

The refinements for parametric judgment and sequents simply delegate to their
subterms as well:
\begin{gather*}
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      \App{\OpNabla{\vec{\sigma}}}{
        \MkBTm{\vec{u}}{}{J}
      }
    }{\SortJdg}
  }{
    \IsWf{\Omega}{\Upsilon,\vec{u}:\vec{\sigma}}{\Gamma}{J}{\SortJdg}
  }\\[6pt]
  \infer{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{
      H\gg A
    }{\SortJdg}
  }{
    \IsWf{\Omega}{\Upsilon}{\Gamma}{H}{\SortTele} &
    \IsWf{\Omega}{\Upsilon}{\Gamma}{A}{\SortType}
  }
\end{gather*}
